---
title: "Homework 3"
author: "Graham Montague"
date: "2025-02-05"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Regular Expression 

## Question 1

```{r, eval=FALSE}
Find expression:  \s{2,}
Replace expression: ,

I found multiple spaces with this expression but still left the spaces between the words. 

Here is what the find and replace expressions do:
\s finds the whitespaces
{2,} finds the two or more times
, replaces with a comma
```

## Question 2
```{r, eval=FALSE}
Find expression:  (\w+), (\w+), (.*)
Replace expression: \2 \1 (\3)

Here is what the find expression specifically does:

(\w+), captures the last name in the first group
(\w+), captures the first name in the second group
(.*) captures all text after the second comma including the college in the third group

Here is what the replacement expression does:

\2 replaces the first name in the replace expression
\1 replaces the last name after the first name
(\3) replaces the college in parentheses
```

## Question 3

```{r, eval=FALSE}
Find expression:  (\d{4}.*?\.mp3)\s+
Replace expression: \1\n

Here is what the find expression does:

\d{4} finds exactly 4 digits in the text
.*? finds any characters using the non-greedy command
\.mp3 finds the the exact writing of '.mp3'
\s+ finds more than or equal to 1 whitespace characters

The replacement expression \1\n puts each matched name to a new line 
```


## Question 4
```{r, eval=FALSE}
Find expression:  (\d{4})\s+(.*?)\.mp3
Replace expression: \1\n

Here is what the find expression does:

(\d{4}) captures the four digit number in the first group
\s+ finds the multiple spaces after the selected number
(.*?) captures the title in the second group and is using the non-greedy command 
\.mp3 finds the writing of '.mp3'

In the replacement expression this is what it does:
  
\2 replaces the title first
_ includes the underscore to the text
\1 replaces the 4-digit number after the underscore above
.mp3 keeps the '.mp3'
```

## Question 5
```{r, eval=FALSE}
Find expression:  ([A-Z])[a-z]+,([a-z]+),\d+\.?\d*,(\d+)
Replace expression: \1_\2,\3

Here is what the find expression does:
  
([A-Z]) captures the first capital letter of the genus text
[a-z]+ finds the rest of the genus but does not capture it
, finds the comma in the text
([a-z]+) captures the species name
, finds the comma
\d+\.?\d* finds the decimal number and does not capture it
, finds the comma
(\d+) captures the final number in the text

Here is what the replacement expression does:
  
\1 replaces the first letter of the genus
_ includes an underscore
\2 replaces the species name
, keeps the comma
\3 replaces the final number
```

## Question 6
```{r, eval=FALSE}
Find expression:  ([A-Z])[a-z]+,([a-z]{4})[a-z]*,\d+\.?\d*,(\d+)
Replace expression: \1_\2,\3

Here is what the find expression does:
  
([A-Z]) captures the first captial letter of the genus
[a-z]+ finds the rest of the genus name, does not capture it
, finds the comma
([a-z]{4}) captures exactly the first 4 letters of the species name
[a-z]* finds remaining letters in the text
, finds the comma
\d+\.?\d* finds the decimal number and does not capture it
, finds the comma
(\d+) captures the final number of the text

Here is what the replacement expression does:

\1 replaces the first letter of the genus
_ includes the underscore
\2 replaces the first 4 letters of the species name
,  the comma remains
\3 the final number remains
```


## Question 7
```{r, eval=FALSE}
Find expression:  ([A-Z][a-z]{2})[a-z]+,([a-z]{3})[a-z]*,(\d+\.\d+),(\d+)
Replace expression: \1\2,\4,\3 


Here is what the expression does:
  
([A-Z][a-z]{2}) captures the the first three letters in the genus
[a-z]+ finds the rest of the genus name
, finds the comma in the text
([a-z]{3}) captures the first 3 letters of the species name
[a-z]* finds and does not capture the rest of the species name
, matches the comma
(\d+\.\d+) captures the decimal number in the text
, matches the comma
(\d+) captures the last number

Here is what the replacement expression does:

\1\2 combines the genus and species
, includes the first comma
\4 replaces the final number so it is seconds
, adds the second comma
\3 finally puts the decimal last
```

## Question 8
Changes made:
First, I want to clean up the pathogen_binary column. The inputs I am expecting to see in this column are 1 and 0s (because it is a binary column). However, there are NA's in the column. Therefore, I want to change these to 0s.

```{r, eval=FALSE}
1. Find expression: NA (I also checked the case sensitive button within the find and replace tab in BBedit)
Replace expression: 0
```

I replaced all of the NAs with 0s. This find expression allowed me to not replace the "na" found in the column name.

Next, I want to clean up bombus_spp and host_plant columns. The bombus_spp includes the species name impatiens that is clearly misspelled throughout the column. In addition, host_plant column also contains special characters that are incorrect.

```{r, eval=FALSE}
2. Find Expressions: [!@#$%^&*()_+-]
Replace expression: with nothing (leave blank)
```

These find and replaces clean up the column to just read impatiens and cleaned up this column. I skipped the header columns so I did not replace those underscores. 

Now I want to add an underscore to every word combo. 

```{r, eval=FALSE}
3. Find Expressions: crown vetch
Replace expression: crown_vetch

Find Expressions: red clover
Replace expression: red_clover

Find Expressions: knot weed
Replace expression: knot_weed

```

After talking with George, there are additional white spaces in the bee_caste column. Therefore, we need to remove the additional white spaces. The find expression finds all of the values in bee_caste column and the white spaces after them. Then the replace statement replaces it with the word and deletes the unnecessary white spaces. To do this see the find and replace expressions below: 

```{r, eval=FALSE}
4. Find Expressions: (worker|male)\s+
Replace expression: \1 (add space after to make sure the worker and 4 are not right next to each other in BBedit if necessary)
```

End of Homework 3










